R for Clinical Study Reports and Submission

### 2021-09-15 project-folder.Rmd
pass spell checking
add placeholder

# Project Folder

In a late-stage clinical trial, the number of A&R deliverables can easily be in the hundreds. 
For an organization, it is possible to have multiple ongoing clinical trials in a clincial program.

A clearly defined project folder structure can have many benefits to 
clinical development teams in an organization. 
Specifically a well defined project folder can achieve:

- Consistency
- Reproducibility
- Automation
- Compliance

## Consistency

For consistency, a well defined folder structure with potential templates
ensures project teams organize A&R work consistently across multiple projects. 
Consistent folder structure also reduce communication cost between study team members 
and enhances the transparency of projects.

For R user, you already benefit from a well defined and consistent folder structure. 
That is the R package folder structure. 
Every R package developer is required to follow the same convention to organize their R
functions, before the function package can be disseminated through the Comprehensive R Archive Network
(CRAN). As a user, you can easily install and user those packages after installing from CRAN.

We recommend to also use R package folder structure to organize analysis scripts 
for clincial trial development. 
Using the R package folder structure to streamline data analysis work has been proposed 
(@marwick2018packaging, @wuanalysis).

In this book, we refer an R package as **project specific R package** 
if the purpose of an R package is to organize analysis scripts for a clinical project.
We refer an R package as **standard R package** if the purpose of an R package is to 
share commonly used R functions to be hosted in an code repository such as CRAN.

Below is a minimal sufficient folders and files based on the R package folder structure.

- `*.Rproj`: RStudio project file used to open RStudio project.
- `DESCRIPTION`: Metadata for a package including authors, license, dependency etc.
- `vignettes/`: Analysis scripts using Rmarkdown.
- `R/`: Project specific R functions.
- `man/`: Manual of project specific R functions.

We demonstrate the the idea using the [`esubdemo`](https://github.com/elong0527/esubdemo) project.

A general discussion of the R package folder structure can be found in Chapter 4 of “R packages:
organize, test, document, and share your code” book (@wickham2015r).

## Reproducibility

Reproducibility of analysis can be challenge. 
By using R package folder structure and proper tools (e.g. `renv`, `packrat`), 
we can easily achieve reproducibility for R and R package versions.

> This is the same level of reproducibility in most SAS environment.
> https://support.sas.com/techsup/pcn/altopsys.html

In `esubdemo` project, a reproducible environment is created when you open the 
`esubdemo.Rproj` from RStudio. 
While RStudio open `esubdemo` project, RStudio will automatically execute R code in `.Rprofile`.
So we are using `.Rprofile` to setup a reproducible environment. 
More details can be found in [https://rstats.wtf/r-startup.html](https://rstats.wtf/r-startup.html).

> `.Rprofile` is only for project specific R package. 
> A standard R package should not use `.Rprofile`.

After the `esubdemo` project is open, the code in `.Rprofile` will automatically check 
current R version is the same as we defined in `.Rprofile`

```{r, eval = FALSE}
R_version <- "4.0.1"	# set up project R version
```

If the R version mismatch, an error message is displayed.

```
Error: The current R version is not the same as the current project in R4.0.1
```

If we are running under the same R version, 
you shall be able to see similar messages to inform how we control R package version.

```
* Project '~/esubdemo' loaded. [renv 0.14.0]
Current oroject R package repository:
    https://mran.microsoft.com/snapshot/2021-08-06
```

The R package version is controlled in two layers. 
Firstly, we define a snapshot date in `.Rprofile`. 
The snapshot date allows to freeze the repository. 
In other words, all R packages installed in this RStudio project 
are based on the frozen R version at the snapshot date. 
Here it is `2021-08-06` by using Microsoft R Application Network (MRAN) server.

```{r, eval = FALSE}
snapshot  <- "2021-08-06" 									                           # set up snapshot date
repos     <- paste0("https://mran.microsoft.com/snapshot/", snapshot)  # set up repository based on snapshot
```

> RStudio Package Manager (RSPM) provided a solution to host both 
> public avaiable and internally developed R packages. 
> However, public RSPM server (https://packagemanager.rstudio.com/client/#/repos/2/overview) 
> did not provide daily snapshot as MRAN had.

Secondly, we use `renv` to lock R package versions and saved in `renv.lock` file. 
The exact same R package versions can be restored with `renv::restore()`.
An introduction of `renv` can be found in its [package website](https://rstudio.github.io/renv/articles/renv.html).

```
* Project '~/github-repo/esubdemo' loaded. [renv 0.14.0]
```

> `renv.lock` file and `renv/` folder are only for project specific R package. 
> A standard R package should not use `renv`.

Once R packages have been properly installed, 
the system will use the R packages located in the search path defined based on the order of `.libPaths()`.
The startup message also provided the R package search path.

```
Below R package path are searching in order to find installed R pacakges in this R session:
    /home/esubdemo/renv/library/R-4.0/x86_64-pc-linux-gnu
    /rtmp/Rtmp8d5mcL/renv-system-library
```

For an organization, a cloud based R environment (e.g. RStudio Workbench) 
can enhance the development and operation of a clinical development team by 
centralize managing R and R package version. 
More details can be found in [https://environments.rstudio.com/](https://environments.rstudio.com/).

In summary, to achieve reproducibility for a project specific R package, 
a clinical project team can work under a controlled R environment 
in the same R version and R package versions defined by a repository snapshot date.

## Automation

- ToDo

There are additional folders and files required in this demo for an analysis project

> people may use different folder name as it is not a standard R package folder.
> These folders and files needs to be added in `.Rbuildignore` to pass
> R package check

- `adam`: ADaM datasets in `.sas7bdat` format.
  + one may also put it in `inst/exdata` following [R package convention](https://r-pkgs.org/data.html)
  + in reality, we suggest to have real data saved outside of this project. (e.g. in a database)
- `output`: TLFs output 
- `renv.lock` and `renv`: R package management using `renv` package. ([Introduction](https://rstudio.github.io/renv/articles/renv.html)) 
- `_pkgdown.yml`: [pkgdown](https://pkgdown.r-lib.org/articles/pkgdown.html) configuration file
- `.Rprofile`: Project startup file to setup running environment including R version, repository, folder path etc. 
  - We further use `inst/startup.R` and `R/zzz.R` to allow the startup file is executed while running. `devtools::load_all()` and RStudio build panel.

## Compliance

- ToDo

>>>>

# Overview

ToDo
